Notes: New gaugings in 3d from three-form

### 2023-05-15 Draft_3dH3.tex
Change of notation for barred indices.

\section{\texorpdfstring{\EE}{E8(8)} ExFT and consistent truncations} \label{s:Review}

\subsection{Review of \texorpdfstring{\EE}{E8(8)} exceptional field theory}
The \EE exceptional field theory, first constructed in ref.~\cite{Hohm:2014fxa}, is an \EE duality-covariant formulation of type II and 11d supergravities. It is defined on a set of $3+248$ coordinates made of three-dimensonal external coordinates $\{x^{\mu}\}$ and internal coordinates $\{Y^{M}\}$ in the 248-dimensional adjoint representation of \EE. The dependance of the fields on these coordinates is constrained by the ``section constraints''\footnote{We use the notation $\otimes$ to indicate that both derivatives may act on different functions.}
\begin{equation} \label{eq:sectionconstraints}
	\begin{cases}
		\eta^{MN}\partial_{M}\otimes\partial_{N} = 0,\\
		f^{MN}{}_{P}\,\partial_{M}\otimes\partial_{N} = 0,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}\partial_{K}\otimes\partial_{L} = 0,
	\end{cases}
\end{equation}
where $f_{MN}{}^{P}$ are the structure constants of \EE, $\eta_{MN} = (1/60)\,f_{MK}{}^{L}f_{NL}{}^{K}$ its Cartan-Killing metric and $(\mathbb{P}_{3875})_{MN}{}^{KL}$ is the projector on the representation $\mathbf{3875}$:
\begin{equation}
	(\mathbb{P}_{3875})_{MN}{}^{KL} = \frac{1}{7}\, \delta_{(M}{}^{K}\delta_{N)}{}^{L}-\frac{1}{56}\,\eta_{MN}\eta^{KL}-\frac{1}{14}\,f^{P}{}_{M}{}^{(K}f_{PN}{}^{L)}.
\end{equation}
Here and in the following, \EE indices are raised and lowered by the Cartan-Killing metric $\eta_{MN}$. The section constraints~\eqref{eq:sectionconstraints} ensure that the fields depends only on the 7- or 8-dimensional physical internal coordinates embeded in $Y^{M}$.

@@

\subsection{Generalized Scherk-Schwarz ansatz and consistent truncations}
The \EE exceptional field theory is well suited to the construction of consistent truncation of type~II and 11d supergravities to three-dimensional maximal supergravity. These truncations arise as generalised Scherk-Schwarz reduction, described by an \EE-valued twist matrix $U_{\fl{M}}{}^{M}$ and a scale factor $\rho$. The compactification Ansätze are as follows~\cite{Galli:2022idq}:
\begin{equation} \label{eq:GSSansatz}
	\begin{aligned}
		g_{\mu\nu}(x,Y)&=\rho(Y)^{-2}\,\mathring{g}_{\mu\nu}(x),\\
		\M_{MN}(x,Y)&=U^{\fl{M}}{}_{M}(Y)\,U^{\fl{N}}{}_{N}(Y)\,\M_{\fl{MN}}(x),\\
		A_{\mu}{}^{M}(x,Y)=&\rho(Y)^{-1}\,U_{\fl{M}}{}^{M}(Y)\,\A_{\mu}{}^{\fl{M}}(x),\\
		B_{\mu\,M}(x,Y)&=\Sigma_{\fl{M}\,M}(Y)\,\A_{\mu}{}^{\fl{M}}(x),
	\end{aligned}
\end{equation}
with
\begin{equation}
	\Sigma_{\fl{M}\,M}=\frac{\rho^{-1}}{60}\,f_{\fl{M}}{}^{\fl{PQ}}\,U_{\fl{P}P}\,\partial_{M}U_{\fl{Q}}{}^{P}.
\end{equation}
The fields with flat indices $\fl{M}$ (in the $\mathbf{248}$ representation of \EE) describe the three-dimensional theory; all the dependence on the internal manifold is factored out in $U_{\fl{M}}{}^{M}$ and $\rho$. Note that with the definition of $\Sigma_{\fl{M}\,M}$ the condition~\eqref{eq:sectionconstraint2} is automatically satisfied.

@@

The constant tensor $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ plays the role of the embedding tensor of the three-dimensional gauged supergravity. It has the following expression:
\begin{equation}
	\begin{aligned}
		X_{\bar M\bar N}{}^{\bar P}&=-\rho^{-1}\,\Gamma_{\bar M\bar N}{}^{\bar P} + \rho^{-1}\,f^{\bar P}{}_{\bar N\bar Q}f^{\bar Q\bar K}{}_{\bar L}\,\Gamma_{\bar K\bar M}{}^{\bar L}-\frac{1}{60}\,\rho^{-1}\,f^{\bar P\bar K}{}_{\bar N}f_{\bar M\bar L}{}^{\bar Q}\,\Gamma_{\bar K\bar Q}{}^{\bar L}\\
		&\quad-\frac{1}{2}\,\rho^{-1}\,f^{\bar P}{}_{\bar N\bar Q}f^{\bar Q\bar K}{}_{\bar M}\,\Gamma_{\bar R\bar K}{}^{\bar R}+\left(\delta_{\bar M}{}^{\bar K}\delta_{\bar N}{}^{\bar P}-\frac{1}{2}\,f_{\bar M}{}^{\bar L\bar K}f_{\bar N\bar L}{}^{\bar P}\right)\,\xi_{\bar K}.
	\end{aligned}
\end{equation}
Here, we defined the ${\rm E}_{8(8)}$ current $\Gamma_{\bar M\bar N}{}^{\bar P}=U_{\bar M}{}^{K}U_{\bar N}{}^{L}\partial_{K}U^{\bar P}{}_{L}$ and the trombone gauging
\begin{equation}
	\xi_{\bar M} = 2\,U_{\bar M}{}^{N}\partial_{N}\rho^{-1}+\rho^{-1}\,\partial_{N}U_{\bar M}{}^{N}.
\end{equation}
In the following, we consider $\xi_{\bar M}=0$. The embedding tensor is most nicely expressed once projected on the adjoint representation
\begin{equation}
	X_{\bar M\bar N} = -\frac{1}{60}\,X_{\bar M\bar P}{}^{\bar Q}\,f_{\bar N\bar Q}{}^{\bar P} = 2\,\rho^{-1}\,\Gamma_{(\bar M\bar N)} - \rho^{-1}\,\Gamma_{\bar K(\bar M}{}^{\bar L}\,f_{\bar N)\bar L}{}^{\bar K},
\end{equation}
with the projection of the current $\Gamma_{\bar M\bar N} = (1/60)\,\Gamma_{\bar M\bar P}{}^{\bar Q}\,f_{\bar N\bar Q}{}^{\bar P}$.

@@

With the generalized Scherk-Schwarz Ansätze~\eqref{eq:GSSansatz}, the action~\eqref{eq:ExFTaction} becomes
\begin{equation}
	\begin{aligned}
		S_{\rm ExFT} \underset{\rm gSS}{=} \int\d^{248}Y\,\rho^{-1}\int \d^{3}x & \bigg[\sqrt{\vert\mathring{g}\vert}\bigg(\mathring{R}+\frac{1}{240}\,{\cal D}_{\mu}\M_{\bar M\bar N}{\cal D}^{\mu}\M^{\bar M\bar N}-V\bigg) \\
		&-\varepsilon^{\mu\nu\rho}\,X_{\bar M\bar N}\,\A_{\mu}{}^{\bar M} \left(\partial_{\nu}\A_{\rho}{}^{\bar N}-\frac{1}{3}\,\llbracket\A_{\nu},\A_{\rho}\rrbracket^{\bar N}\right)\bigg],
	\end{aligned}
\end{equation}
with the potential\footnote{To obtain this potential, one has to add the total derivative $-\partial_{M}\left(\sqrt{\vert g\vert}\,\partial_{N}\M^{MN}+(4/3)\,\M^{MN}\partial_{N}\sqrt{\vert g\vert}\right)$ to $\sqrt{\vert g\vert}\,\mathscr{L}_{\rm int}$.}
\begin{equation}
	V = X_{\bar M\bar N}\,X_{\bar P\bar Q}\left(\frac{1}{28}\,\M^{\bar M\bar P}\M^{\bar N\bar Q}+\frac{1}{2}\,\M^{\bar M\bar P}\eta^{\bar N\bar Q}+\frac{13}{28}\,\eta^{\bar M\bar P}\eta^{\bar N\bar Q}\right).
\end{equation}

>>>>

% \section{\texorpdfstring{\EE}{E8(8)} ExFT and consistent truncations}

\section{Review of \texorpdfstring{\EE}{E8(8)} exceptional field theory} \label{s:Review}

The \EE exceptional field theory, first constructed in ref.~\cite{Hohm:2014fxa}, is an \EE duality-covariant formulation of type II and 11d supergravities. It is defined on a set of $3+248$ coordinates made of three-dimensonal external coordinates $\{x^{\mu}\}$ and internal coordinates $\{Y^{M}\}$ in the 248-dimensional adjoint representation of \EE. The dependance of the fields on these coordinates is constrained by the ``section constraints''\footnote{We use the notation $\otimes$ to indicate that both derivatives may act on different functions.}
\begin{equation} \label{eq:sectionconstraints}
	\begin{cases}
		\eta^{MN}\partial_{M}\otimes\partial_{N} = 0,\\
		f^{MN}{}_{P}\,\partial_{M}\otimes\partial_{N} = 0,\\
		(\mathbb{P}_{3875})_{MN}{}^{KL}\partial_{K}\otimes\partial_{L} = 0,
	\end{cases}
\end{equation}
where $f_{MN}{}^{P}$ are the structure constants of \EE, $\eta_{MN} = (1/60)\,f_{MK}{}^{L}f_{NL}{}^{K}$ its Cartan-Killing metric and $(\mathbb{P}_{3875})_{MN}{}^{KL}$ is the projector on the representation $\mathbf{3875}$:
\begin{equation}
	(\mathbb{P}_{3875})_{MN}{}^{KL} = \frac{1}{7}\, \delta_{(M}{}^{K}\delta_{N)}{}^{L}-\frac{1}{56}\,\eta_{MN}\eta^{KL}-\frac{1}{14}\,f^{P}{}_{M}{}^{(K}f_{PN}{}^{L)}.
\end{equation}
Here and in the following, \EE indices are raised and lowered by the Cartan-Killing metric $\eta_{MN}$. The section constraints~\eqref{eq:sectionconstraints} ensure that the fields depends only on the 7- or 8-dimensional physical internal coordinates embeded in $Y^{M}$.

@@

\section{Consistent truncations to 3-dimensional \texorpdfstring{${\cal N}=16$}{N=16} gauged supergravity}
The \EE exceptional field theory is well suited to the construction of consistent truncation of type~II and 11d supergravities to three-dimensional maximal supergravity. These truncations arise as generalised Scherk-Schwarz reductions, described by an \EE-valued twist matrix $U_{\fl{M}}{}^{M}$ and a scale factor $\rho$. The compactification Ansätze are as follows~\cite{Galli:2022idq}:
\begin{equation} \label{eq:GSSansatz}
	\begin{aligned}
		g_{\mu\nu}(x,Y)&=\rho(Y)^{-2}\,\mathring{g}_{\mu\nu}(x),\\
		\M_{MN}(x,Y)&=U^{\fl{M}}{}_{M}(Y)\,U^{\fl{N}}{}_{N}(Y)\,\M_{\fl{MN}}(x),\\
		A_{\mu}{}^{M}(x,Y)=&\rho(Y)^{-1}\,U_{\fl{M}}{}^{M}(Y)\,\A_{\mu}{}^{\fl{M}}(x),\\
		B_{\mu\,M}(x,Y)&=\Sigma_{\fl{M}\,M}(Y)\,\A_{\mu}{}^{\fl{M}}(x),
	\end{aligned}
\end{equation}
with
\begin{equation}
	\Sigma_{\fl{M}\,M}=\frac{\rho^{-1}}{60}\,f_{\fl{M}}{}^{\fl{PQ}}\,U_{\fl{P}P}\,\partial_{M}U_{\fl{Q}}{}^{P}.
\end{equation}
The fields with flat indices $\fl{M}$ (in the $\mathbf{248}$ representation of \EE) describe the three-dimensional theory; all the dependence on the internal manifold is factored out in $U_{\fl{M}}{}^{M}$ and $\rho$. Note that with the definition of $\Sigma_{\fl{M}\,M}$ the condition~\eqref{eq:sectionconstraint2} is automatically satisfied.

@@

The constant tensor $X_{\fl{M}\fl{N}}{}^{\fl{P}}$ plays the role of the embedding tensor of the three-dimensional gauged supergravity. It has the following expression:
\begin{equation}
	\begin{aligned}
		X_{\fl{MN}}{}^{\fl{P}}&=-\rho^{-1}\,\Gamma_{\fl{MN}}{}^{\fl{P}} + \rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}f^{\fl{QK}}{}_{\fl{L}}\,\Gamma_{\fl{KM}}{}^{\fl{L}}-\frac{1}{60}\,\rho^{-1}\,f^{\fl{PK}}{}_{\fl{N}}f_{\fl{ML}}{}^{\fl{Q}}\,\Gamma_{\fl{KQ}}{}^{\fl{L}}\\
		&\quad-\frac{1}{2}\,\rho^{-1}\,f^{\fl{P}}{}_{\fl{NQ}}f^{\fl{QK}}{}_{\fl{M}}\,\Gamma_{\fl{RK}}{}^{\fl{R}}+\left(\delta_{\fl{M}}{}^{\fl{K}}\delta_{\fl{N}}{}^{\fl{P}}-\frac{1}{2}\,f_{\fl{M}}{}^{\fl{LK}}f_{\fl{NL}}{}^{\fl{P}}\right)\,\xi_{\fl{K}}.
	\end{aligned}
\end{equation}
Here, we defined the ${\rm E}_{8(8)}$ current $\Gamma_{\fl{MN}}{}^{\fl{P}}=U_{\fl{M}}{}^{K}U_{\fl{N}}{}^{L}\partial_{K}U^{\fl{P}}{}_{L}$ and the trombone gauging
\begin{equation}
	\xi_{\fl{M}} = 2\,U_{\fl{M}}{}^{N}\partial_{N}\rho^{-1}+\rho^{-1}\,\partial_{N}U_{\fl{M}}{}^{N}.
\end{equation}
In the following, we consider $\xi_{\fl{M}}=0$. The embedding tensor is most nicely expressed once projected on the adjoint representation
\begin{equation}
	X_{\fl{MN}} = -\frac{1}{60}\,X_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}} = 2\,\rho^{-1}\,\Gamma_{(\fl{MN})} - \rho^{-1}\,\Gamma_{\fl{K}(\fl{M}}{}^{\fl{L}}\,f_{\fl{N})\fl{L}}{}^{\fl{K}},
\end{equation}
with the projection of the current $\Gamma_{\fl{MN}} = (1/60)\,\Gamma_{\fl{MP}}{}^{\fl{Q}}\,f_{\fl{NQ}}{}^{\fl{P}}$.

@@

With the generalized Scherk-Schwarz Ansätze~\eqref{eq:GSSansatz}, the action~\eqref{eq:ExFTaction} becomes
\begin{equation}
	\begin{aligned}
		S_{\rm ExFT} \underset{\rm gSS}{=} \int\d^{248}Y\,\rho^{-1}\int \d^{3}x & \bigg[\sqrt{\vert\mathring{g}\vert}\bigg(\mathring{R}+\frac{1}{240}\,{\cal D}_{\mu}\M_{\fl{MN}}{\cal D}^{\mu}\M^{\fl{MN}}-V\bigg) \\
		&-\varepsilon^{\mu\nu\rho}\,X_{\fl{MN}}\,\A_{\mu}{}^{\fl{M}} \left(\partial_{\nu}\A_{\rho}{}^{\fl{N}}-\frac{1}{3}\,\llbracket\A_{\nu},\A_{\rho}\rrbracket^{\fl{N}}\right)\bigg],
	\end{aligned}
\end{equation}
with the potential\footnote{To obtain this potential, one has to add the total derivative $-\partial_{M}\left(\sqrt{\vert g\vert}\,\partial_{N}\M^{MN}+(4/3)\,\M^{MN}\partial_{N}\sqrt{\vert g\vert}\right)$ to $\sqrt{\vert g\vert}\,\mathscr{L}_{\rm int}$.}
\begin{equation}
	V = X_{\fl{MN}}\,X_{\fl{PQ}}\left(\frac{1}{28}\,\M^{\fl{MP}}\M^{\fl{NQ}}+\frac{1}{2}\,\M^{\fl{MP}}\eta^{\fl{NQ}}+\frac{13}{28}\,\eta^{\fl{MP}}\eta^{\fl{NQ}}\right).
\end{equation}
